Quantitative Benchmarks for Theory Exploration

### 2017-04-19 PhDSymposium2017/abstract/abstract.tex
Split up abstract from slides

\documentclass[]{default}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{amssymb,amsmath}
\usepackage{paralist}
\usepackage{tikz-qtree}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Exploring Software with Symbolic and Statistical Algorithms},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}
\newcommand{\feature}[1]{\phi(#1)}
\newcommand{\id}[1]{\texttt{"#1"}}
\newcommand{\CVar}{\texttt{Var}}
\newcommand{\CLit}{\texttt{Lit}}
\newcommand{\CApp}{\texttt{App}}
\newcommand{\CLam}{\texttt{Lam}}
\newcommand{\CLet}{\texttt{Let}}
\newcommand{\CCase}{\texttt{Case}}
\newcommand{\CType}{\texttt{Type}}
\newcommand{\CLocal}{\texttt{Local}}
\newcommand{\CGlobal}{\texttt{Global}}
\newcommand{\CConstructor}{\texttt{Constructor}}
\newcommand{\CLitNum}{\texttt{LitNum}}
\newcommand{\CLitStr}{\texttt{LitStr}}
\newcommand{\CAlt}{\texttt{Alt}}
\newcommand{\CDataAlt}{\texttt{DataAlt}}
\newcommand{\CLitAlt}{\texttt{LitAlt}}
\newcommand{\CDefault}{\texttt{Default}}
\newcommand{\CNonRec}{\texttt{NonRec}}
\newcommand{\CRec}{\texttt{Rec}}
\newcommand{\CBind}{\texttt{Bind}}

\title{Exploring Software with Symbolic and Statistical Algorithms}
  \author{
          Chris Warburton \\
      University of Dundee \\
      \texttt{cmwarburton@dundee.ac.uk}
    }
\date{}

\begin{document}
\maketitle
\begin{abstract}
We aim to aid programmers in understanding and maintaining software
by increasing the capability of \emph{theory exploration} systems for
discovering novel, unexpected properties of programs and libraries.
Existing theory exploration systems find equivalent program fragments via brute-force
search, which doesn't scale well to realistic codebases. We narrow down
the scope of each exploration run using statistics derived from the
semantic similarity of the source code being explored, using a process
called \emph{recurrent clustering}.
\end{abstract}

@@

As computers become ever more important in research, industry and
society, so does the ability to program them effectively, economically
and \emph{correctly}. Whilst tools exist to help programmers understand
and manipulate software, ranging from linters \cite{Johnson78lint} through to formal
verification systems \cite{boyer1983proof}, these are often limited in scope (e.g. to syntax rather than semantics) or require precise guidance from the user. One
reason is that automating the search for patches \cite{Forrest.Nguyen.Weimer.ea:2009}, refactorings,
optimisations \cite{phothilimthana2016scaling}, proofs \cite{rosen2012proving}, and other routine programming tasks is
prohibitively expensive due to the size of the search space.

@@

To reduce the user's need to cherry-pick signatures, our approach
accepts a large signature (e.g. a complete Haskell package) and uses
statistical machine learning methods to select smaller signatures
automatically, using a \emph{recurrent clustering} method inspired by
those of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and \textsc{ACL2(ml)} \cite{Heras.Komendantskaya.Johansson.ea:2013}.

Recurrent clustering attempts to cluster expressions based on their
Abstract Syntax Trees (ASTs), an example of which is shown in Figure \ref{fig:astexample}. First the recursively-structured ASTs are
transformed into vectors of fixed length, by truncating the tree
structure and listing the node labels in breadth-first order.

@@

\begin{figure}
    \begin{small}
      \begin{tikzpicture}[sibling distance=0pt]
        \tikzset{sibling distance=0pt}
        \tikzset{level distance=20pt}
        \Tree[ .$\CLam$
                  $\id{a}$
                  [ .$\CCase$
                       [ .$\CVar$
                            [ .$\CLocal$
                                 $\id{a}$ ]]
                       $\id{b}$
                       [ .$\CAlt$
                            $\CDataAlt$
                            [ .$\CVar$
                                 $\CConstructor$ ]]
                       [ .$\CAlt$
                            $\CDataAlt$
                            [ .$\CApp$
                                 [ .$\CVar$
                                      [ .$\CGlobal$
                                           $\id{even}$ ]]
                                 [ .$\CVar$
                                      [ .$\CLocal$
                                           $\id{n}$ ]]]
                            $\id{n}$ ]]]
      \end{tikzpicture}
    \end{small}
    \caption{AST for the \texttt{odd} function. The variable names \texttt{a}, \texttt{b}, etc. are chosen arbitrarily.}
    \label{fig:astexample}
\end{figure}

@@

Figure \ref{fig:astexample} shows the AST for the following Haskell function \texttt{odd}, which determines if a Peano numeral is odd:

@@

The most difficult aspect of performing this exploration was obtaining real code in a usable format, due to the myriad edge-cases encountered. We make extensive use of Haskell's existing infrastructure, including the \textsc{GHC} compiler, the \textsc{cabal} build system, the \textsc{Hackage} code repository and the \textsc{Nix} package manager. Unfortunately, some of these systems are rather monolithic, which makes it difficult to invoke particular algorithms, such as \textsc{GHC}'s type class resolution, on their own. Whilst we have worked around these issues, e.g. using meta-programming, this introduces unnecessary latency, fragility and complexity.

@@

\bibliographystyle{plain}
\bibliography{../Bibtex}

>>>>

\documentclass[]{default}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{amssymb,amsmath}
\usepackage{paralist}
\usepackage{tikz-qtree}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Quantitative Benchmarks for Theory Exploration},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}
\newcommand{\feature}[1]{\phi(#1)}
\newcommand{\id}[1]{\texttt{"#1"}}
\newcommand{\CVar}{\texttt{Var}}
\newcommand{\CLit}{\texttt{Lit}}
\newcommand{\CApp}{\texttt{App}}
\newcommand{\CLam}{\texttt{Lam}}
\newcommand{\CLet}{\texttt{Let}}
\newcommand{\CCase}{\texttt{Case}}
\newcommand{\CType}{\texttt{Type}}
\newcommand{\CLocal}{\texttt{Local}}
\newcommand{\CGlobal}{\texttt{Global}}
\newcommand{\CConstructor}{\texttt{Constructor}}
\newcommand{\CLitNum}{\texttt{LitNum}}
\newcommand{\CLitStr}{\texttt{LitStr}}
\newcommand{\CAlt}{\texttt{Alt}}
\newcommand{\CDataAlt}{\texttt{DataAlt}}
\newcommand{\CLitAlt}{\texttt{LitAlt}}
\newcommand{\CDefault}{\texttt{Default}}
\newcommand{\CNonRec}{\texttt{NonRec}}
\newcommand{\CRec}{\texttt{Rec}}
\newcommand{\CBind}{\texttt{Bind}}

\title{Quantitative Benchmarks for Theory Exploration}
  \author{Chris Warburton \\
          University of Dundee \\
          \texttt{cmwarburton@dundee.ac.uk}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
  \emph{Automated theory exploration} has the potential to help scientists,
  mathematicians and programmers discover unexpected properties of their models,
  theories and programs; but has so far been limited to small-scale experiments,
  with hand-picked data and qualitative results. We address these problems by
  introducing a novel benchmark suite, orders of magnitude larger than previous
  data sets, and use it to evaluate a variety of theory exploration approaches
  (both existing and novel) and measure how they scale. It is our hope that the
  availability of a robust method for comparing systems will encourage
  competition and innovation in this field, as it has done for e.g.\ automated
  theorem proving.
\end{abstract}

@@

Many fields represent knowledge using \emph{formal systems}, for example
scientific models, mathematical theories and computer programs. As requirements
grow, these systems can be modified in two ways: becoming more complicated, and
becoming more numerous. Complication makes a system's behaviour harder to
predict, whilst proliferation dilutes the attention that can be given to each
particular system. This impacts aspects like safety and security, for example
proving that a program will never leak confidential information, or that a model
of some mechanism will never enter an unsafe configuration.

One way to mitigate these problems is through automation: \emph{automated theory
exploration} is the task of discovering interesting properties and behaviours of
a given formal system, such as a software library or set of equations. Such
tools can be used by domain experts, to discover structure, redundancy and
patterns in a system; and, more importantly, to be \emph{surprised} by
unexpected insights.

\begin{figure}
  \begin{equation*}
    plus(zero,    y) = y \\
    plus(succ(x), y) = succ(plus(x, y)) \\
    length(nil)        = zero \\
    length(cons(x, y)) = succ(length(y)) \\
    append(nil,       z) = z \\
    append(cons(x,y), z) = cons(x, append(y, z)) \\
    map(f, nil)        = nil \\
    map(f, cons(x, y)) = cons(f(x), map(f, y))
  \end{equation*}
  \caption{Functions.}
  \label{fig:defs}
\end{figure}

Our running example will be the software library\footnote{Our treatment of
  software is directly applicable to mathematics \cite{wadler2015propositions},
  and is also applicable to scientific models if we encode them as programs
  which output predicted observations \cite{davies1990physical}
} in figure~\ref{fig:defs}

functions for manipulating data. This

We can think of this as an axiomatic equational theory; it's also We will concentrate on the example of a software library, but this is

The problem of \emph{automated theorem proving} (ATP) was one of the first tasks
studied in the field of artificial intelligence \cite{newell1956logic,
  sutcliffe2001evaluating}: given some statement about a formal system, prove
that it is true. Many ATP approaches have been devised, and systems are compared
by evaluating their performance on large benchmark problems, such as TPTP
(Thousands of Problems for Theorem Provers).

In contrast, theory \emph{exploration}progress in ATP has been through progTheory theorem proving a the  experts with a system for  discovering the behaviour of such systems has impact may  can exhibit Both present challenges to understandingare As these systems become more complexcomputers become ever more important in research, industry and
society, so does the ability to program them effectively, economically
and \emph{correctly}. Whilst tools exist to help programmers understand
and manipulate software, ranging from linters \cite{Johnson78lint} through to
formal verification systems \cite{boyer1983proof}, these are often limited in
scope (e.g.\ to syntax rather than semantics) or require precise guidance from
the user. One reason is that automating the search for patches
\cite{Forrest.Nguyen.Weimer.ea:2009}, refactorings, optimisations
\cite{phothilimthana2016scaling}, proofs \cite{rosen2012proving}, and other
routine programming tasks is prohibitively expensive due to the size of the
search space.

@@

To reduce the user's need to cherry-pick signatures, our approach
accepts a large signature (e.g.\ a complete Haskell package) and uses
statistical machine learning methods to select smaller signatures
automatically, using a \emph{recurrent clustering} method inspired by
those of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and \textsc{ACL2(ml)} \cite{Heras.Komendantskaya.Johansson.ea:2013}.

Recurrent clustering attempts to cluster expressions based on their
Abstract Syntax Trees (ASTs), an example of which is shown in Figure~\ref{fig:astexample}. First the recursively-structured ASTs are
transformed into vectors of fixed length, by truncating the tree
structure and listing the node labels in breadth-first order.

@@

\begin{figure}
    \begin{small}
      \begin{tikzpicture}[sibling distance=0pt]
        \tikzset{sibling distance=0pt}
        \tikzset{level distance=20pt}
        \Tree[ .$\CLam$
                  $\id{a}$
                  [ .$\CCase$
                       [ .$\CVar$
                            [ .$\CLocal$
                                 $\id{a}$ ]]
                       $\id{b}$
                       [ .$\CAlt$
                            $\CDataAlt$
                            [ .$\CVar$
                                 $\CConstructor$ ]]
                       [ .$\CAlt$
                            $\CDataAlt$
                            [ .$\CApp$
                                 [ .$\CVar$
                                      [ .$\CGlobal$
                                           $\id{even}$ ]]
                                 [ .$\CVar$
                                      [ .$\CLocal$
                                           $\id{n}$ ]]]
                            $\id{n}$ ]]]
      \end{tikzpicture}
    \end{small}
    \caption{AST for the \texttt{odd} function. The variable names \texttt{a}, \texttt{b}, etc.\ are chosen arbitrarily.}
    \label{fig:astexample}
\end{figure}

@@

Figure~\ref{fig:astexample} shows the AST for the following Haskell function \texttt{odd}, which determines if a Peano numeral is odd:

@@

The most difficult aspect of performing this exploration was obtaining real code in a usable format, due to the myriad edge-cases encountered. We make extensive use of Haskell's existing infrastructure, including the \textsc{GHC} compiler, the \textsc{cabal} build system, the \textsc{Hackage} code repository and the \textsc{Nix} package manager. Unfortunately, some of these systems are rather monolithic, which makes it difficult to invoke particular algorithms, such as \textsc{GHC}'s type class resolution, on their own. Whilst we have worked around these issues, e.g.\ using meta-programming, this introduces unnecessary latency, fragility and complexity.

@@

\bibliographystyle{plain}
\bibliography{./Bibtex}
